The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free e-learning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2009-08-12 script/de/chapters/cryptomethods.tex
CT script, cover with smaller size; many small corrections; consistency for all chapters like bibliography everywhere in PDF navigation and link moves to correct page (de+en)
CT script, chap 01: added section 'Ciphers and cryptanalysis for educational purposes' and according Sage sample (en)
CT script: added some references especially about Sage (en)

\hypertarget{Kapitel_1}{}
\chapter{Verschlüsselungsverfahren}
(Bernhard Esslinger, Jörg-Cornelius Schneider, Mai 1999; Updates: Dez. 2001, Feb. 2003,
Juni 2005, Juli 2007, Juli 2009)

@@

% --------------------------------------------------------------------------
% \section{Verschlüsselung}

@@

\begin{sloppypar}
Einen sehr guten Überblick zu den Verfahren bietet das Buch von Bruce
Schneier \cite{cm:Schneier1996}. Grundsätz"-lich unterscheidet man zwischen
symmetrischen und asymmetrischen Verfahren zur Verschlüsse"-lung.
\end{sloppypar}
% --------------------------------------------------------------------------
\section[Symmetrische Verschlüsselung]
{Symmetrische Verschlüsselung\footnotemark}
   \footnotetext{% XXX dieses Prozent ist nötig, sonst startet die Fußnote mit einem Blank !
     Mit CrypTool\index{CrypTool} können Sie über das Menü
     {\bf Ver-/Entschlüsseln \textbackslash{} Symmetrisch (modern)} folgende 
     modernen symmetrischen Verschlüsselungsverfahren ausführen: \\
     IDEA, RC2, RC4, DES (ECB), DES (CBC), Triple-DES (ECB), Triple-DES (CBC),
     MARS (AES-Kandidat), RC6 (AES-Kandidat), Serpent (AES-Kandidat), 
     Twofish (AES-Kandidat), Rijndael (offizielles AES-Verfahren).
     }
\nopagebreak
Bei der {\em symmetrischen} Verschlüsselung 
\index{Verschlüsselung!symmetrisch} müssen Sender und Empfänger
über einen gemeinsamen (geheimen) Schlüssel verfügen, den sie vor
Beginn der eigentlichen Kommunikation ausgetauscht haben. Der Sender
benutzt diesen Schlüssel, um die Nachricht zu verschlüsseln und der
Empfänger, um diese zu entschlüsseln.\par \vskip + 3pt

@@

Hier werden die Daten mittels symmetrischer Verfahren
verschlüsselt: der Schlüssel ist ein vom Absender zufällig\footnote{%
  Die Erzeugung zufälliger Zahlen ist ein wichtiger Bestandteil 
  kryptographisch sicherer Verfahren. Mit CrypTool\index{CrypTool} 
  können Sie über das Menü {\bf Einzelverfahren \textbackslash{}
  Zufallsdaten erzeugen} verschiedene Zufallszahlengeneratoren ausprobieren.
  Über das Menü {\bf Analyse \textbackslash{} Zufallsanalyse} können Sie
  verschiedene Testverfahren für Zufallsdaten auf binäre Dokumente
  anwenden. \\
  Bisher konzentriert sich CrypTool auf kryptographisch starke 
  Pseudozufallszahlengeneratoren. Nur im Secude-Generator
  \index{SECUDE IT Security GmbH} wird eine "`echte"'
  Zufallsquelle einbezogen. 
}\index{Zufall}
generierter Sitzungsschlüssel (session key)\index{Session Key}, der
nur für diese Nachricht verwendet wird. Anschließend wird dieser
Sitzungsschlüssel mit Hilfe des asymmetrischen Verfahrens
verschlüsselt und zusammen mit der Nachricht an den Empfänger
übertragen. Der Empfänger kann den Sitzungsschlüssel mit Hilfe
seines geheimen Schlüssels bestimmen und mit diesem dann die
Nachricht entschlüsseln. Auf diese Weise nutzt man das bequeme
Schlüsselmanagement \index{Schlüsselmanagement} asymmetrischer
Verfahren und kann trotzdem große Datenmengen schnell und effektiv mit
symmetrischen Verfahren verschlüsseln.

>>>>

\hypertarget{Kapitel_1}{}
\chapter{Verschlüsselungsverfahren}
\label{Label_Kapitel_1}
(Bernhard Esslinger, Jörg-Cornelius Schneider, Mai 1999; Updates: Dez. 2001, Feb. 2003,
Juni 2005, Juli 2007, August 2009)

@@

@@

\begin{sloppypar}
Einen sehr guten Überblick zu den Verfahren bietet das Buch von Bruce
Schneier \cite{cm:Schneier1996}. Grundsätz"-lich unterscheidet man zwischen
symmetrischen und asymmetrischen Verfahren zur Verschlüsse"-lung.
\end{sloppypar}

% --------------------------------------------------------------------------
\section[Symmetrische Verschlüsselung]
{Symmetrische Verschlüsselung\footnotemark}
   \footnotetext{% XXX dieses Prozent ist nötig, sonst startet die Fußnote mit einem Blank !
     Mit CrypTool\index{CrypTool} können Sie über das Menü
     {\bf Ver-/Entschlüsseln \textbackslash{} Symmetrisch (modern)} folgende 
     modernen symmetrischen Verschlüsselungsverfahren ausführen: \\
     IDEA, RC2, RC4, DES (ECB), DES (CBC), Triple-DES (ECB), Triple-DES (CBC),
     MARS (AES-Kandidat), RC6 (AES-Kandidat), Serpent (AES-Kandidat), 
     Twofish (AES-Kandidat), Rijndael (offizielles AES-Verfahren).
     }
\nopagebreak
Bei der {\em symmetrischen} Verschlüsselung 
\index{Verschlüsselung!symmetrisch} müssen Sender und Empfänger
über einen gemeinsamen (geheimen) Schlüssel verfügen, den sie vor
Beginn der eigentlichen Kommunikation ausgetauscht haben. Der Sender
benutzt diesen Schlüssel, um die Nachricht zu verschlüsseln und der
Empfänger, um diese zu entschlüsseln.\par \vskip + 3pt

@@

Hier werden die Mengen-Daten mittels symmetrischer Verfahren
verschlüsselt: Der Schlüssel ist ein vom Absender zufällig\footnote{%
  Die Erzeugung zufälliger Zahlen ist ein wichtiger Bestandteil 
  kryptographisch sicherer Verfahren. Mit CrypTool\index{CrypTool} 
  können Sie über das Menü {\bf Einzelverfahren \textbackslash{}
  Zufallsdaten erzeugen} verschiedene Zufallszahlengeneratoren ausprobieren.
  Über das Menü {\bf Analyse \textbackslash{} Zufallsanalyse} können Sie
  verschiedene Testverfahren für Zufallsdaten auf binäre Dokumente
  anwenden. \\
  Bisher konzentriert sich CrypTool auf kryptographisch starke 
  Pseudozufallszahlengeneratoren. Nur im Secude-Generator
  \index{SECUDE IT Security GmbH} wird eine "`echte"'
  Zufallsquelle einbezogen. 
}\index{Zufall}
generierter geheimer Sitzungsschlüssel (session key)\index{Session Key},
der nur für diese Nachricht verwendet wird.

Anschließend wird dieser Sitzungsschlüssel mit Hilfe des asymmetrischen
Verfahrens verschlüsselt und zusammen mit der Nachricht an den Empfänger
übertragen.

Der Empfänger kann den Sitzungsschlüssel mit Hilfe seines geheimen
Schlüssels bestimmen und mit diesem dann die Nachricht entschlüsseln.

Auf diese Weise profitiert man von dem bequemen Schlüsselmanagement\index{Schlüsselmanagement}
asymmetrischer Verfahren (mit öffentlichem und privatem Schlüssel), und
man profitiert von der Schnelligkeit symmetrischer Verfahren, um große
Datenmengen zu verschlüsseln (mit den geheimen Schlüsseln).

### 2009-08-12 script/de/chapters/menus.tex
CT script: added some references especially about Sage (en)

   % Eyecatcher_Neue-CrypTool-Version
Dieser Anhang enthält auf der folgenden Seite den kompletten Menübaum der
CrypTool\index{CrypTool}-Version 1.4.30\footnote{%
  Von den Zukunftsversionen CrypTool 2 und JCrypTool im CrypTool-Projekt gibt
  es bisher nur (stabile) Betaversionen. Wenn es hier Releaseversionen gibt,
  werden im Skript die entsprechen Menüpfade und Menübäume ergänzt.
}.

>>>>

   % Eyecatcher_Neue-CrypTool-Version
Dieser Anhang enthält auf der folgenden Seite den kompletten Menübaum der
CrypTool\index{CrypTool}-Version 1.4.30\footnote{%
  Parallel zu CrypTool 1.x\index{CrypTool 1.x} werden im CrypTool-Projekt
  momentan auch die Zukunftsversionen CrypTool 2\index{CrypTool 2.0} und
  JCrypTool\index{JCrypTool 1.0} entwickelt.\\
  Diese Zukunftsversionen sind zur Zeit (August 2009) noch Betaversionen,
  diese sind aber schon stabil genug um von Endbenutzern genutzt werden
  zu können. Wenn es hier Releaseversionen gibt, werden im Skript die
  entsprechenden Menüpfade und Menübäume ergänzt.
}.

### 2009-08-12 script/en/chapters/cryptomethods.tex
CT script: added some references especially about Sage (en)

In this case the data is encrypted using symmetric procedures: The key is a
session key\index{Session key} generated by the sender
randomly\footnote{%
An important part of cryptographically secure techniques is to generate 
random numbers. Within CrypTool\index{CrypTool} you can check out
different random number generators using the menu path
{\bf Indiv. Procedures \textbackslash{} Generate Random Numbers}. 
Using the menu path {\bf Analysis \textbackslash{} Analyze Randomness}
you can apply different test methods for random data to binary documents. \\
Up to now CrypTool has concentrated on cryptographically strong 
pseudo random number generators. Only the integrated Secude\index{SECUDE IT Security}
generator involves a "pure" random source. 
}\index{Random}
that is only used for this message.
This session key is encrypted using the asymmetric procedure, and then
transmitted to the recipient together with the message. Recipients can determine
the session key using their secret keys and then use the session key to encrypt
the message. In this way, we can benefit from the easy key management
\index{Key management} of asymmetric procedures and encrypt large quantities of data
quickly and efficiently using symmetric procedures.

@@

So some simplified variants of modern symmetric ciphers like DES, IDEA or AES
were developed for educational purposes in order to allow beginners to perform
encryption and decryption by hand and gain a better understanding of how the algorithms
work in detail. One is e.g. S-AES (Simplified-AES) by Prof. Ed Schaefer and his students%
\footnote{
    \url{http://math.scu.edu/~eschaefe/getfile.pdf} \\
    \url{http://www.scu.edu/cas/research/cryptography.cfm}
}. Another one is Mini-AES (see chapter~\ref{CM_Sage_Mini-AES} ``\nameref{CM_Sage_Mini-AES}'').

These also help to understand and apply the according cryptanalysis methods.

\begin{itemize}

@@

The following Sage code from the release note of Sage 4.1\\
\url{http://mvngu.wordpress.com/2009/07/12/sage-4-1-released/}\\
calls the implementation of the Mini Advanced Encryption Standard.

The Sage module \texttt{crypto/block\_cipher/miniaes.py} supports Mini-AES to allow students
to explore the working of a modern block cipher. This is a simplified variant of
the Advanced Encryption Standard (AES) to be used for cryptography education.
Mini-AES is described at \cite{cm:Phan2002}.

@@

\bibitem[Hoffman2006]{cm:Hoffman2006} \index{Hoffman 2006} 
       Nick Hoffman, \\
       {\em A SIMPLIFIED IDEA ALGORITHM},\\
       \url{http://www.nku.edu/~christensen/simplified%20IDEA%20algorithm.pdf}

\bibitem[Lucks-DuD2002]{cm:Lucks-DuD2002}  \index{Lucks 2002}
       Stefan Lucks, R\"udiger Weis, \\
       {\em Neue Ergebnisse zur Sicherheit des Verschl\"usselungsstandards AES}, 
       in DuD Dec. 2002.

@@

\bibitem[Robshaw2002a]{cm:Robshaw2002a}  \index{Robshaw 2002}
       S.P. Murphy, M.J.B. Robshaw, \\
       {\em Essential Algebraic Structure within the AES}, 
       June 5, 2002, Crypto 2002,  \\
       \href{http://www.isg.rhul.ac.uk/\~{}mrobshaw/rijndael/rijndael.html}
       {\texttt{http://www.isg.rhul.ac.uk/~mrobshaw/rijndael/rijndael.html}}

\bibitem[Robshaw2002b]{cm:Robshaw2002b}  \index{Robshaw 2002}
       S.P. Murphy, M.J.B. Robshaw, \\
       {\em Comments on the Security of the AES and the XSL Technique}, 
       September 26, 2002, \\
       \href{http://www.isg.rhul.ac.uk/\~{}mrobshaw/rijndael/rijndael.html}
       {\texttt{http://www.isg.rhul.ac.uk/~mrobshaw/rijndael/rijndael.html}}

@@

\bibitem[Wobst-iX2002]{cm:Wobst-iX2002}  \index{Wobst 2002}
       Reinhard Wobst, \\
       {\em Angekratzt - Kryptoanalyse von AES schreitet voran}, 
       in iX Dec. 2002, \\
       plus the reader's remark by Johannes Merkle in iX Feb. 2003.

>>>>

In this case the bulk data is encrypted using symmetric procedures: The key
used for this is a secret session key\index{Session key} generated by the
sender randomly\footnote{%
An important part of cryptographically secure techniques is to generate 
random numbers. Within CrypTool\index{CrypTool} you can check out
different random number generators using the menu path
{\bf Indiv. Procedures \textbackslash{} Generate Random Numbers}. 
Using the menu path {\bf Analysis \textbackslash{} Analyze Randomness}
you can apply different test methods for random data to binary documents. \\
Up to now CrypTool has concentrated on cryptographically strong 
pseudo random number generators. Only the integrated Secude\index{SECUDE IT Security}
generator involves a "pure" random source. 
}\index{Random}
that is only used for this message.

This session key is then encrypted using the asymmetric procedure, and
transmitted to the recipient together with the message.

Recipients can determine the session key using their private keys and
then use the session key to encrypt the message.

In this way, we can benefit from the easy key management\index{Key management}
of asymmetric procedures (using public/private keys) and we benefit from the
efficiency of symmetric procedures to encrypt large quantities of data
(using secret keys).

@@

So some simplified variants of modern symmetric ciphers like DES, IDEA or AES
were developed for educational purposes in order to allow beginners to perform
encryption and decryption by hand and gain a better understanding of how the
algorithms work in detail.
These also help to understand and apply the according cryptanalysis methods.

One of these variants is e.g. S-AES (Simplified-AES) by Prof. Ed Schaefer
and his students \cite{cm:Musa-etal2003}%
\footnote{
    See the articles
    ``Devising a Better Way to Teach and Learn the Advanced Encryption Standard''\\
    at the university's news at\\
    \url{http://math.scu.edu/~eschaefe/getfile.pdf} \\
    \url{http://www.scu.edu/cas/research/cryptography.cfm}
}.
Another one is Mini-AES \cite{cm:Phan2002}
(see chapter~\ref{CM_Sage_Mini-AES} ``\nameref{CM_Sage_Mini-AES}''):
\begin{itemize}

@@

The Sage module \texttt{crypto/block\_cipher/miniaes.py} supports Mini-AES to allow
students to explore the working of a modern block cipher.

Mini-AES, originally described at \cite{cm:Phan2002}, is a simplified variant of the Advanced
Encryption Standard (AES) to be used for cryptography education.

How to use Mini-AES is exhaustively described at the Sage reference page\\
\url{http://www.sagemath.org/doc/reference/sage/crypto/block_cipher/miniaes.html}.

The following Sage code from the release note of Sage 4.1 calls the implementation
of the Mini Advanced Encryption Standard.\\
See \url{http://mvngu.wordpress.com/2009/07/12/sage-4-1-released/}.

@@

\bibitem[Hoffman2006]{cm:Hoffman2006} \index{Hoffman 2006} 
       Nick Hoffman, \\
	2006 \\
       {\em A SIMPLIFIED IDEA ALGORITHM},\\
       \url{http://www.nku.edu/~christensen/simplified%20IDEA%20algorithm.pdf}

\bibitem[Lucks-DuD2002]{cm:Lucks-DuD2002}  \index{Lucks 2002}
       Stefan Lucks, R\"udiger Weis, \\
       {\em Neue Ergebnisse zur Sicherheit des Verschl\"usselungsstandards AES},\\
       In: DuD, Dec 2002.

@@

\bibitem[Robshaw2002a]{cm:Robshaw2002a}  \index{Robshaw 2002}
       S.P. Murphy, M.J.B. Robshaw, \\
       {\em Essential Algebraic Structure within the AES}, \\
       June 5, 2002, Crypto 2002,  \\
       \href{http://www.isg.rhul.ac.uk/\~{}mrobshaw/rijndael/rijndael.html}
       {\texttt{http://www.isg.rhul.ac.uk/~mrobshaw/rijndael/rijndael.html}}

\bibitem[Robshaw2002b]{cm:Robshaw2002b}  \index{Robshaw 2002}
       S.P. Murphy, M.J.B. Robshaw, \\
       {\em Comments on the Security of the AES and the XSL Technique}, \\
       September 26, 2002, \\
       \href{http://www.isg.rhul.ac.uk/\~{}mrobshaw/rijndael/rijndael.html}
       {\texttt{http://www.isg.rhul.ac.uk/~mrobshaw/rijndael/rijndael.html}}

@@

\bibitem[Wobst-iX2002]{cm:Wobst-iX2002}  \index{Wobst 2002}
       Reinhard Wobst, \\
       {\em Angekratzt - Kryptoanalyse von AES schreitet voran}, \\
       In: iX, Dec 2002; \\
       plus the reader's remark by Johannes Merkle in iX Feb 2003.

### 2009-08-12 script/en/chapters/menus.tex
CT script: added some references especially about Sage (en)

This appendix contains at the following page the complete menu tree of
CrypTool\index{CrypTool} version 1.4.30\footnote{%
  The future versions CrypTool 2 and JCrypTool of the CrypTool project are
  currently there only as (stable) beta versions. If the according release
  versions are available, we will add the appropriate menus and menu trees.\\
}.

>>>>

This appendix contains at the following page the complete menu tree of
CrypTool\index{CrypTool} version 1.4.30\footnote{%
  In parallel to CrypTool 1.x\index{CrypTool 1.x} the future versions
  CrypTool 2\index{CrypTool 2.0} and JCrypTool\index{JCrypTool 1.0}
  are currently developed in the CrypTool project.\\
  These future versions are currently (August 2009) beta versions, but they
  are stable enough to be used by end-users too. If the according release
  versions are available, we will add the appropriate menus and menu trees.\\
}.

### 2009-08-12 script/en/chapters/what-is-sage.tex
CT script: added some references especially about Sage (en)

This script includes numerous code samples using Sage. Sage is an open
source computer algebra system (CAS) that supports teaching, study and
research in mathematics.  It combines many high-quality open
source packages and provides access to their functionalities via a
common interface, namely, a Python based programming language.  Sage
can be used as a powerful desktop calculator, as a tool to help
(undergraduate) students study mathematics, or as a programming
environment for prototyping algorithms and research in algorithmic
aspects of mathematics.

With respect to studying cryptography, Sage modules in the directory
\url{SAGE_ROOT/devel/sage-main/sage/crypto} can be used to complement
a first course in cryptography.  David Kohel's notes at
\begin{center}
  \url{http://www.sagemath.org/library/crypto.pdf}
\end{center}
can be used to teach such a course that incorporates Sage.

@@

>>>>

This script includes numerous code samples using Sage. Sage is an open
source computer algebra system (CAS) that supports teaching, study and
research in mathematics.  It combines many high-quality open
source packages and provides access to their functionalities via a
common interface, namely, a Python\footnote{%
There is also an easy interface to the C language, called Cython, which can be used
to speed up own functions in Sage very much.\\
See \url{http://openwetware.org/wiki/Open_writing_projects/Sage_and_cython_a_brief_introduction}.
} based programming language.

Sage can be used as a powerful desktop calculator, as a tool to help
(undergraduate) students study mathematics, or as a programming
environment for prototyping algorithms and research in algorithmic
aspects of mathematics.

\noindent You can get an easy impression e.g. with David Joyner's ``Invitation to Sage''
(last updated 2009):\\
{\centering \url{http://sage.math.washington.edu/home/wdj/teaching/calc1-sage/an-invitation-to-sage.pdf}.}

\noindent The official Sage online documentation\footnote{%
  The according official PDF documents can be downloaded at\\
  \url{http://www.sagemath.org/help.html}
} is available at: \url{http://www.sagemath.org/doc}.

%% \centering statt \begin{center} ... \end{center}, da dies zusätzlich vertical space erzeugt.
There are lots of bigger PDF or HTML documentation about using Sage,
so we name only a few of them:
A good starting point is the ``Library'' from the Sage main page:\\
{\centering \url{http://www.sagemath.org/library/index.html}}
and the ``Documentation'' Project\\
{\centering \url{http://wiki.sagemath.org/DocumentationProject}}.

These also contain teaching material at\\
{\centering \url{http://wiki.sagemath.org/Teaching_with_SAGE}}.

With respect to studying cryptography, Sage modules in the directory
\url{SAGE_ROOT/devel/sage-main/sage/crypto} can be used to complement
a first course in cryptography\footnotetext{%
- Overview, what crypto currently is in Sage:\\
~ ~ \url{http://www.sagemath.org/doc/reference/sage/crypto/}\\
- Discussions about the teaching related aspect of development crypto in Sage:\\
~ ~ \url{http://groups.google.com/group/sage-devel/browse_thread/thread/c5572c4d8d42d081}
}.

Especially David Kohel's notes at\\
{\centering \url{http://www.sagemath.org/library/crypto.pdf} }
or the same eventually newer at\\
{\centering \url{http://sage.math.washington.edu/home/wdj/teaching/kohel-crypto.pdf} }\\
can be used to teach such a course that incorporates Sage.

@@

xxxxxxxxxxx
Sage tutorial about Programming
"Loading and Attaching SAGE files"
\url{http://www.sagemath.org/doc/tutorial/programming.html#loading-and-attaching-sage-files}
xxxxxxxxxxxxx
